Intersection Type Algebras

### 2025-03-13 Relations/Paper/Implications.tex

The implications in this diagram are formalized in \texttt{ARS-Implications.agda}. The classical property of being minimally decidable is 
required to complete the implication from $\SN$ to $\WN$, as can be seen in our formalization \sacomment{Rewrite this section}

@@

The isMinDec property \sacomment{refer to list of classical properties here} states that all elements either have a relation to some other element, or 
there is no other element to which they have a relation. \greyout{If the element relates to no other element then it is terminating 
and so trivially $\WN$. If the element does relate to some other element $y \in A$, then we recursively apply our claim to 
$y$ and because we know that we have the property of $\SN$ we must eventually reach a terminating element.}

>>>>

The implications in this diagram are formalized in \texttt{ARS-Implications.agda}. 
The implication from $\SN$ to $\WN$ relies on the classical decidability of the property 
of being $R$-minimal, as demonstrated in our formalization:

@@

### 2025-03-13 Relations/Paper/WFFigure.tex

\begin{figure}[h]
    \centering
    \begin{tikzpicture}[node distance=2cm, auto]
      % Standard definitions
      \node (std1) [draw] {isWFacc};
      \node (std2) [right of=std1, draw] {isWFind};
      \node (std3) [right of=std2, draw] {isWFseq};
      \node (std4) [right of=std3, draw] {isWFmin};

>>>>

\begin{figure}[h]
    \centering
    \begin{tikzpicture}[node distance=2cm, auto]
      % Standard definitions
      \node (std1) [draw] {isWFacc};
      \node (std2) [right of=std1, draw, xshift=1cm] {isWFind};
      \node (std3) [right of=std2, draw, xshift=1cm] {isWFseq};
      \node (std4) [right of=std3, draw, xshift=1cm] {isWFmin};

### 2025-03-13 Relations/Paper/Well-foundedness.tex

\section{Well-foundedness}
\label{sec:Well-foundedness}

>>>>

\section{Well-foundedness}

\newcommand{\isWFacc}{\mathtt{isWFacc}}
\newcommand{\isWFseq}{\mathtt{isWFseq}}
\newcommand{\isWFmin}{\mathtt{isWFmin}}
\newcommand{\isWFaccm}{\mathtt{isWFacc-}}
\newcommand{\isWFseqm}{\mathtt{isWFseq-}}
\newcommand{\isWFminm}{\mathtt{isWFmin-}}

\newcommand{\isMinDec}{\mathtt{isMinDec}}

\newcommand{\then}{\Longrightarrow}
\label{sec:Well-foundedness}

An abstract reduction $\to_R$ is strongly normalizing
if and only if the converse relation is well-founded.
In this section, we will compare several definitions of well-foundedness.

\subsection{The main definitions}

The various formulations of well-foundedness are based on the following notions.
\begin{definition}
  Let $R \subseteq A \times A$.  We write $Rxy$ when $(x,y) \in R$.
  \begin{enumerate}
    \item $P \subseteq A$ is \emph{inductive}
    if, for each $x \in A$, $(\forall y. Ryx \to y \in P)$ implies $x \in P$.

    \item $x \in A$ is \emph{accessible} if $x \in P$ for each inductive $P$.

    \item $x \in A$ is \emph{$P$-minimal} if $x \in P$ and for all $y$,
    $Ryx$ implies $y \notin P$.

    \item $s : \nat \to A$ is \emph{$R$-decreasing} if for all $k$, $(s(k+1),s(k)) \in R$.
  \end{enumerate}
\end{definition}

In \cite{Terese}, well-foundedness is defined as in the classical mathematics textbooks,
as the absence of an infinite $R$-decreasing sequence.
Translating this statement directly into type theory yields the formula
\[\tag{$\isWFseqm$} \lnot \big(\exists s : \nat \to A \forall k : \nat. (s(k+1),s(k)) \in R \big) \]
(The label $\isWFseqm$ refers to the name of the corresponding term in \texttt{Wellfoundedness.agda}.)
Constructively, this is not a useful hypothesis, as it does not provide any positive computational content and requires an explicit sequence to be defined.

A stronger formulation asserts that, for any sequence, there exists an element
in the sequence that does not reduce to the next one.  This yields the statement
\[\tag{$\isWFseq$} \forall s : \nat \to A \exists k : \nat. (s(k),s(k+1)) \notin R \]

These two conditions are classically equivalent, but not constructively so.
Indeed, $\isWFseq$ certainly implies $\isWFseqm$, but the converse only holds classically.

Another classical formulation of well-foundedness
asserts that every non-empty subset
contains a minimal element.  That is,
\[ \tag{$\isWFmin$} \forall P \subseteq A. P \neq \emptyset \to \exists x \in P. (\forall y. Ryx \to y\notin P) \]

While classically equivalent to the previous one, see \cite[Exercise A.1.8]{Terese},
constructively we only have the implication $\isWFmin \then \isWFseq$.

The canonical constructive definition of well-foundedness is that
every inductive predicate is universally true.
This is constructively equivalent to the statement that every element
is accessible.  In Agda, these concepts are encoded as follows.
{\small
\begin{verbatim}
  _-inductive_ : 𝓟 A → Set
  _-inductive_ φ = ∀ x → (∀ y → R y x → φ y) → φ x

  _isWFind : Set₁
  _isWFind = ∀ (φ : 𝓟 A) → _-inductive_ φ → ∀ x → φ x

  data _-accessible : 𝓟 A where
    acc : ∀ {x : A} → (∀ y → R y x → _-accessible y) → _-accessible x

  _isWFacc : Set
  _isWFacc = ∀ (x : A) → x ∈ _-accessible
\end{verbatim}
}
Although the two notions are equivalent, the latter definition allows us to do induction on the proof of accessibility of an individual element $x \in A$.  This makes such proofs easier to write in Agda, thus we take
\verb|_isWFacc| as \emph{the} standard notion of well-foundedness for relations.

\subsection{Relations between the main definitions}

We have now seen five notions of a well-founded relation, only two of which
are seen to be logically equivalent.

This raises several questions.
\begin{itemize}
\item Are all these notions truly distinct?  If one cannot furnish a proof of implication,
can we provide a counterexample?
\item Under what hypotheses do the implications become provable?
\item How ``admissible'' are such hypotheses, and how often can we hope for them to be valid?
\end{itemize}

The situation of separating these notions is complicated by the fact that, for all
\emph{definable} predicates, the classical and constructive notions do coincide.
(We have not proved this, but see \cite{Berardi} for a related result.)
So there is no hope to discriminate these notions via explicit counterexamples.
Rather, we could attempt to show that one does not imply the other by proving that
such an implication would yield a ``constructive taboo'': a proposition known to be
unprovable without additional classical assumptions.

For example, the archetype of a well-founded relation is the strict order \verb|<|
of natural numbers.  The usual induction readily validates $\isWFacc$ for \verb|<|.
  At the same time, the file \texttt{WFCounters.agda} shows that
asserting $\isWFmin$ for \verb|<| implies that every subset of the natural numbers is decidable.  Since this is clearly not provable in Agda without classical axioms,
this argument shows that the implication from $\isWFacc$ to $\isWFmin$ is not provable.

Such distinctions have consequences for rewriting theory.
For example, $x\in \SN_R$ iff $x$ is $\bar{R}$-accessible (where $\bar{R}$ denotes
the converse of $R$).  Thus, if $\bar{R}$ is well-founded, $R \models \gSN$.
Yet actually computing the normal form of a given $x$ seems to require a
strong form of decidability of being $R$-minimal, namely:
\[
\tag{$\isMinDec$} \forall x. \left(\exists y. Ryx\right) \lor \left(\forall y. \lnot Ryx\right)
\]
(Note that asking whether $x$ is a normal form constitutes the weaker disjunction
$\lnot\lnot\left(\exists y. Ryx\right) \lor \left(\forall y. \lnot Ryx\right)$.)

On the other hand, $\isWFmin$ trivially implies that each $x \in A$ is weakly normalizing.
Indeed, given $x \in A$, consider the predicate $P_x(y) = x \mstep y$.
This predicate is non-empty, since the empty reduction yields a proof of $P_x(x)$.
Let $n$ be a $P_x$-minimal element.  Then $x \mstep n$.  Moreover, for any $y$,
If $x \rstep y$, then $\lnot(P_x)$.  But since reductions from $x$ are closed downward,
no such $y$ can exist, and hence $n$ is a normal form of $x$.

% \greyout{From the perspective of rewriting theory, it would seem wild indeed if
% $\SN$ did not imply $\WN$.
%
These questions motivated us to take a closer look at the issue of well-foundedness.
Our analysis revealed a number of additional related hypotheses, having a complex
network of relations.  %These are displayed in Figure \ref{f:wf}.
These are recorded in the file \verb|Wellfounded.agda|

### 2025-03-13 UTA/notes.tex
more ura

@@

\section*{01.17}
TODO:
\begin{itemize}
  \item Need to start writing the paper!
  \item Need a pretty-printer/helper functions for using the Agda code.
  \item Need to refactor/cleanup the code
  \item Work on the conjecture
  \begin{enumerate}
    \item One direction should be easy
    \item Is the other direction even true?
  \end{enumerate}
  \item Generalizing the code to preorders/intersection types
  \item Verify correctness of the code: prove it computes the most general unifier
  equivalent to the starting list of constraints
  \item Implementing the decision algorithm for the generated recursive type algebra
  \item How to implement/define the ``free'' ita/meet semilattices in type theory
  --- related to: How to define strict order between types.
  \item Implement/verify the decider for the subtyping relation between trees.
\end{itemize}

@@

>>>>

\newcommand{\mcI}{\mathcal{I}}

@@

\section*{01.17}
TODO:
\begin{itemize}
  \item Need to start writing the paper!
  \item Need a pretty-printer/helper functions for using the Agda code.
  \item Need to refactor/cleanup the code
  \item Work on the conjecture
  \begin{enumerate}
    \item One direction should be easy
    \item Is the other direction even true?
  \end{enumerate}
  \item Generalizing the code to preorders/intersection types
  \item Verify correctness of the code: prove it computes the most general unifier
  equivalent to the starting list of constraints
  \item Implementing the decision algorithm for the generated recursive type algebra
  \item How to implement/define the ``free'' ita/meet semilattices in type theory
  --- related to: How to define strict order between types.
  \item Implement/verify the decider for the subtyping relation between trees.
  \item To look into strong normalization criterion.
\end{itemize}

@@

\section{Conjecture}

Let $\types$ be the set of simple types.
An \emph{inequational type theory} is a set of type inequalities
  \[ \mcI = \setof{A_i \le B_i \mid i \in I, A_i \in \types, B_i \in \types} \]

In most cases, we will be interested in the case when $I = |\mcI|$ is a finite set.

Let $n \in \nat$, $I = \setof{1,..,n}$, and $\mcI$ an inequational type theory
over $I$.

Let $V$ be the set of type atoms --- to be treated as type variables ---
occuring in $\mcI$.

The theory $\mcI$ can be transformed into a set $\mcI^+,\mcI^-$, where
each $(A \le B) \in \mcI^+$ is of the form $\alpha \le T$,
for some $\alpha \in V$ and $T \in \types$, and
each $(A \le B) \in \mcI^-$ is of the form $T \le \alpha$,
for some $\alpha \in V$ and $T \in \types$.

For $\alpha \in V$, put
\begin{align*}
\mcI^+(\alpha) &= \setof{A \mid (\alpha \le A) \in \mcI^+}\\
\mcI^-(\alpha) &= \setof{A \mid (A \le \alpha) \in \mcI^-}
\end{align*}

We can further transform this subtyping theory by replacing the set of variables
$V$ by $V^+ \cup V^-$, where
\begin{align*}
  V^+ &= \setof{\alpha^+ \mid \alpha \in V} \\
  V^- &= \setof{\alpha^- \mid \alpha \in V}
\end{align*}

For $\alpha \in V$, and for $A=A(\alpha) \in \types$, let $A(\alpha^-,\alpha^+)$
be obtained by replacing all positive occurrences of $\alpha$ in $A$ with $\alpha^+$,
and all negative occurrences of $\alpha$ in $A$ with $\alpha^-$.

Let $\mcS^0 = \setof{\alpha^- \le \alpha^+ \mid \alpha \in V}$.

Let $\mcS(\mcI) = \mcS^0 \cup \mcS^+ \cup \mcS^-$, where
\begin{align*}
  \mcS^+ = \setof{\alpha^+ \le A(\alpha^+,\alpha^-) \mid A \in \mcI^+(\alpha)}\\
  \mcS^- = \setof{A(\alpha^-,\alpha^+) \le \alpha^- \mid A \in \mcI^-(\alpha)}
\end{align*}

By taking inversion closure, these sets can be assumed to be minimal.

\textsc{Proposition.}  $\mcI$ admits a solution in $\types_\cap$ if and only if
$\mcS(\mcI)$ admits a solution in $\types_\cap$.

\textsc{Conjecture.} $\mcS(\mcI)$ admits a solution in $\types_\cap$ if and only if
there is no ``circular constraint'': a type $B$ such that, for some
$\alpha \in V$, $A^+ \in \mcI^+(\alpha)$, $A^- \in \mcI^-(\alpha)$, one has:
% \[ \greyout{A^+(\alpha^-,\alpha^+) \le B \le A^-(\alpha^-,\alpha^+)} \]
\[ A^+(\alpha^+,\alpha^-) \le B \le A^-(\alpha^-,\alpha^+) \]

\subsection*{2024.06.11 -- Reasons for believing the conjecture}

The condition is obviously necessary, since a circular constraint immediately
results in $\alpha = A(\alpha)$, which is not solvable in $\types_\cap$.

As for the necessity, we can make the following observations.

First, the ``positive'' side, $\mcS^+$ can be solved by taking
$\alpha^+ := \alpha \cap A_1 \cap \cdots \cap A_k$, where $\mcI^+(\alpha) = \setof{A_1,\dots,A_k}$.

As for the negative side, we could invert the constraints by mapping everything into
a new type atom, $\gamma$.

But how do we get the most ``general'' unifier?  That is, how do we get to the situation
where, we know that the constraints are solvable, but want the most generic scheme
of the solution types?
